---
title: "Phytobiomes_hemp-microbiome-analysis"
author: "Allison Rietman"
date: "2024-10-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Code Overview

This document contains all code for Rietman et al. 2024 Phytobiomes manuscript which investigates the effects of fungicide application and pathogen inoculation on field grown hemp AND a biological control trial which challenges corn earworm larvae with diet inoculated with an isolated fungal endophyte recovered from hemp leaves. 

Fungicide-Pathogen Inoculation Trial
1. Endophyte community analysis
2. Diversity metric analysis
3. Indicator species analysis
4. Bacterial abundance analysis
5. Visual presence of pathogen infection analysis

Biological Control Trial
1. Choice trial analyses
2. Survival trial analysis


## Load packages and data 
```{r pkgs}
library(ggplot2)
library(ggpubr)
library(vegan)
library(car)
library(lme4)
library(viridis)
library(dplyr)
library(tidyverse)
library(reshape2)
library(spaa)
library(indicspecies)
library(emmeans)
library(multcomp)
library(plotrix)
library(survival)
library(survminer)
```

## Here, we are uploading all data for subsequent analyses and adjusting structure of dataset for easy use downstream
```{r read data}
pivot_data <-read.csv("otu_matrix_S1_seaquencher.csv",
                      header = TRUE, row.names = 1) 

env <-read.csv("new_env_S21_fungicide.csv", header=TRUE, row.names=1, sep=",")
  env$plot.no<-as.factor(env$plot.no)
  env$treatment<-as.factor(env$treatment)
  
LMA.data <- read.csv("plant-growth-metrics-S21.csv", header = TRUE)
  LMA.data$plant.height.cm <- as.numeric(LMA.data$plant.height.cm)
  LMA.data$plot.no<-as.factor(LMA.data$plot.no)
  LMA.data$sample.id<-as.factor(LMA.data$sample.id)
  LMA.data$treatment<-as.factor(LMA.data$treatment)
  
new.bac <- read.csv("new_bacteria_abun_S21.csv")
  new.bac$treatment<-as.factor(new.bac$treatment)
  new.bac$plot.no<-as.factor(new.bac$plot.no)
  new.bac$sample.id<-as.factor(new.bac$sample.id)

choice <- read.csv("ChoiceTrialData_CornEarwormOR.csv", header = TRUE, sep = ",")
  choice$SubjectID<-as.factor(choice$SubjectID)
  choice$choice<-as.factor(choice$choice)
  choice$day<-as.factor(choice$day)
  
food <- read.csv("ChoiceTrialData_FoodMass.csv", header = TRUE, sep = ",")
  food$subjectID<-as.factor(food$subjectID)
  food$foodtype<-as.factor(food$foodtype)
  # Add in a variable for difference in starting and beginning food mass
  food$food.diff <- (food$initialfoodmass - food$finalfoodmass)
  
levels <- read.csv("LevelsTrialData_CornEarwormOR.csv", header = TRUE, sep = ",")
  levels$SubjectID<-as.factor(levels$SubjectID)
  levels$Trial<-as.factor(levels$Trial)
  levels$Day<-as.factor(levels$Day)
  levels$Dead<-as.factor(levels$Dead)
  
corncox <- read.csv("cox_cornearnworm.csv", header = TRUE, sep = ",")
  # NOTE - status column indicates if dead, 1 = dead, 0 = pupated
  
```

## ENDOPHYTE COMMUNITY ANALYSIS 
Here, our data undergoes numerous steps and filtering before the results of the analysis. We first remove singletons from our dataset, the perform a hellinger correction, then we make the ordination of our community data

```{r community analysis}

colSums(pivot_data) # get the sum of fungal presence in each column

data_no_sing<-pivot_data[colSums(pivot_data)>1 ] # only include columns that have sums greater than 1, 
                                                 # i.e., no singletons
colSums(data_no_sing) # get the sum of fungal presence in each column W/O singletons
rowSums(data_no_sing) # sum of fungi in each leaf

# make data with row sums (fungi in each leaf) 
data_row_sum <- data_no_sing[ rowSums(data_no_sing)!=0, ] 
    
# get rid of rows (samples) that don't have any fungi present
# Check to see if all rows have values greater than 0
sort(rowSums(data_row_sum))
str(data_row_sum)

# Use pivot table w/o blank rows to run data
# Use hellinger method
data_hel <- decostand(data_row_sum, method = "hellinger")
data_hel # check to see if values are less than 1

# make ordination
ord<-metaMDS(data_hel, trymax=20,k=2) # must have vegan loaded
ord
ord$points

```

Now, we are using our environmental data and matching it with our transformed data from above that omits singletons. 

```{r}

# remove rows (ie. samples) that do not have any colonization from env data
env2<-env[row.names(env) %in% row.names(data_row_sum),]
env2$Treatment<-as.factor(env2$treatment)
str(env2)


# Determine how many plots represent the treatment groups in adjusted data
env2 %>% count(treatment)

```

We plot the results from the ordination. We included two plots 1) the plot which excludes extreme points and 2) one that includes all points. We chose to use the plot that excludes the extreme points so that the data could be visualized. We denote this in our paper. 

```{r echo=TRUE}
## use below code to get rid of "outlier" plots 106-1, 303-1, 203-1; I changed axis limits

plot(ord, type = "n", main="treatment", ylim = c(-1,1), xlim = c(-3, 3))
  # adjusted axis limits again to center the ordination on the plot 
#plot(ord, type = "n", main="treatment")
points(ord, display = "sites", cex = 1.5, pch=21, col="black", bg=env2$Treatment)
```

```{r echo=TRUE}
## OR use this code to show all points

plot(ord, type = "n", main="all points, treatment")
  # adjusted axis limits again to center the ordination on the plot 
plot(ord, type = "n", main="treatment")
points(ord, display = "sites", cex = 1.5, pch=21, col="black", bg=env2$Treatment) 
ordiellipse(ord, env2$Treatment, label= TRUE, col="black", kind=c("sd"))
    # putting ellipses around to show treatment group clustering
ordilabel(ord, display = "sites")
    # label with sites for diagnostic purposes; not for publication 

```

After all of that, we run our permutation on the community data

```{r permutation of community data}

# Does treatment have an effect on endo community? 
adonis2(data_hel ~ env2$Treatment, permutations = 999, method = "bray")

# Calculate Bray-Curtis distances between samples
dis <- vegdist(data_hel)

## Calculate multivariate dispersions for each treatment
mod <- betadisper(dis, env2$Treatment)
mod

```

Now, let's look at the spread of the different groups 
 permutation-based test of multivariate homogeneity of group dispersions (variances); looks not at composition differences between groups, but differences in homogeneity

```{r echo=TRUE}

# Permutation test for F
pmod <- permutest(mod, permutations = 99, pairwise = FALSE)
pmod
```

```{r echo=TRUE}
# Tukey's Honest Significant Differences
(mod.HSD <- TukeyHSD(mod))
plot(mod.HSD)

  # Together these results suggest that the groups differ significantly in their dispersion (spread)
  # Fungicide group has more spread than natural or inoculation group

```

## DIVERSITY METRIC ANALYSES OF ENDOPHYTE COMMUNITY

In this section, we calculate Shannon diversity and species richness.

First, let's tackle Shannon diversity

```{r shannon diversity}

d <- diversity(pivot_data, index = "shannon")
shannon <- as.data.frame(d)
shannon$Treatment <- env$treatment
shannon$Height <- env$plant.height.cm
shannon$Plot<-env$plot.no

shannon

# Calculating average Shannon D for each treatment group
d_natural<-shannon[which(shannon$Treatment == "natural"),]
d_innoculation<-shannon[which(shannon$Treatment == "innoculation"),]
d_fungicide<-shannon[which(shannon$Treatment == "fungicide"),]

mean(d_natural$d)
mean(d_innoculation$d)
mean(d_fungicide$d)

median(d_natural$d)
median(d_innoculation$d)
median(d_fungicide$d)
```

```{r echo=TRUE}
# Run ANOVA on Shannon diversity 
anova_diversity2 <- lm(d ~ Treatment, data = shannon)
    # this code would get rid of collinearity with Treatment +  Plot
                    
Anova(anova_diversity2, type="III") 
```

```{r echo=TRUE}
# Test assumptions and plot residuals 
res_div2 <- resid(anova_diversity2)
plot(fitted(anova_diversity2), res_div2)
abline(0,0)

# Run a levene's test
library(car)
leveneTest(d ~ Treatment, data = shannon)
```

Now, let's move on to species richness

```{r species richness}
# For this, we use the package "vegan"

# How speciose are my communities? (summation of number of species)
sppr <- specnumber(pivot_data)

# Let's extract the rows of interest form the env dataframe
treat.df <- data.frame(env$plot.no, env$treatment)

# Make a plot of species count and treatment 
plot(sppr ~ treat.df$env.treatment)
```

```{r echo=TRUE}
# Run the test
sppr_aov <- aov(sppr ~ env.treatment, data = treat.df)
summary(sppr_aov)
```

```{r echo=TRUE}
# Test the assumptions and plot residuals
# Plot the residuals 
res_rich <- resid(sppr_aov)
plot(fitted(sppr_aov), res_rich)
abline(0,0)
```

```{r echo=TRUE}
# Run a levene's test
leveneTest(sppr ~ env.treatment, data = treat.df)
```

```{r}
# Calculate mean species richness
treat.df$sppr <- sppr # add the values of sppr to the treat.df data frame

r_natural<-treat.df[which(treat.df$env.treatment == "natural"),]
r_innoculation<-treat.df[which(treat.df$env.treatment == "innoculation"),]
r_fungicide<-treat.df[which(treat.df$env.treatment == "fungicide"),]

mean(r_natural$sppr)
mean(r_innoculation$sppr)
mean(r_fungicide$sppr)


# Now, let's perform a Tukey's test
tukey.rich <- TukeyHSD(sppr_aov)
tukey.rich
plot(tukey.rich)
```

## INDICATOR SPECIES ANALYSIS 
For this, we use the package "indicspecies" (already loaded)

```{r indicator species, echo=TRUE}
# Let's determine indicator species

indval3 <- multipatt(pivot_data, env$treatment, 
                    control = how(nperm = 9999))
  # NOTE - help pages suggest to make one data frame that includes treatment
  # and pivot data, this doesn't always work! Better to use pivot data and a separate
  # file for treatment groups 
  # https://cran.r-project.org/web/packages/indicspecies/vignettes/IndicatorSpeciesAnalysis.html#indicato r-species-analysis-using-multipatt

summary(indval3, indvalcomp=TRUE)
```

```{r echo=TRUE}
# Let's look at more species that aren't significant (a = 1)
# This is just for fun!!! 
summary(indval3, alpha = 1)
```

## ANALYZING PLANT GROWTH METRIC DATA 
Here, we look at plant height and leaf mass per area (LMA) 

```{r plant growth metric}
lm_lma<-lm(LMA.g.m2 ~ treatment, data = LMA.data)
lm_height<-lm(plant.height.cm ~ treatment, data = LMA.data)

Anova(lm_lma)
Anova(lm_height)
```

```{r}
# Let's check assumptions 
# Plot the residuals for height
res_height <- resid(lm_height)
plot(fitted(lm_height), res_height)
abline(0,0)

# Run a levene's test for height
library(car)
leveneTest(plant.height.cm ~ treatment, data = LMA.data)

# Plot the residuals for lma
res_lma <- resid(lm_lma)
plot(fitted(lm_lma), res_lma)
abline(0,0)

# Run a levene's test for lma
leveneTest(LMA.g.m2 ~ treatment, data = LMA.data)
```

## DETERMINING DIFFERENCES IN BACTERIAL ABUNDANCE 

```{r bac abundance}
# make linear model
lm.bac.new <- lm(total.bac ~ treatment, data = new.bac)
anova(lm.bac.new)


plot(total.bac ~ treatment, data = new.bac)
```

```{r}
# Get the mean abundance of each treatment group 
bac_natural<-new.bac[which(new.bac$treatment == "natural"),]
bac_innoculation<-new.bac[which(new.bac$treatment == "innoculation"),]
bac_fungicide<-new.bac[which(new.bac$treatment == "fungicide"),]

# Run Tukey's HSD test
bac.aov <- aov(lm.bac.new) # need to use aov function to run Tukey's, I think
tukey.bac <- TukeyHSD(bac.aov)
tukey.bac
plot(tukey.bac)
```

### VISIBLE PATHOGEN DAMAGE ANALYSIS 

```{r path damage analysis}
# We need to create a subset of the env data set for this analysis 
sep<-env
sep$septoria<-pivot_data$Septoria_dysentericae
sep
```

Some notes about this dataset for clarity: the column "final.pathogen" represents the visual appearance of pathogen damage tabulated from the observers. "septoria" represents the presence of Septoria dysentericae isolates found in culture. 

```{r}
##Looking at the 0/1 binomial response data if pathogen damage was there or not
##Take-home point: Fungicide-treated plants had less visible damage
str(sep)
lm_damage <- glm((final.pathogen ~ treatment), family = binomial, data = sep)
Anova(lm_damage)
summary(lm_damage)

##Can't run a Tukey directly on a GLM with binomial response, have to calculate predicted probabilities and then run it
# Compute estimated marginal means
library(emmeans)
library(multcomp)
emm <- emmeans(lm_damage, ~treatment)
# Perform post hoc tests for pairwise comparisons
posthoc <- contrast(emm, method = "pairwise", adjust = "tukey")

# View the results
summary(posthoc)
```

# BIOLOGICAL CONTROL TRIAL 
Now, we get to shift gears and move onto the second part of the manuscript which is the corn earworm biological control trial. 

## CHOICE TRIAL ANALYSIS
```{r choice analysis, echo=TRUE}
# Create linear model to run ANOVA 
lm_foodpref <- lm(food.diff~foodtype, data = food)

#Testing normality of residuals using Shapiro-Wilk test
shapiro.test(lm_foodpref$resid)
# p = 0.53 so we can assume normality

# Plot to test for all other assumptions
par(mfrow = c(2,2))
plot(lm_foodpref)
```

```{r echo=TRUE}
anova(lm_foodpref)
```

```{r}
# Determine mean food consumption of each treatment group
eaten_control <- food[which(food$foodtype == "no"),]
eaten_high <- food[which(food$foodtype == "high"),]

# Now, we calculate mean amount of food eaten in each group (g) 
mean(eaten_control$food.diff)
mean(eaten_high$food.diff)

# Now, let's calculate standard error
print(std.error(eaten_control$food.diff))
print(std.error(eaten_high$food.diff))
```

Here, we run an anova as a "good measure" to see if gave the larvae the same amount of food (in grams) at the beginning of the experiment. 
```{r}
## Just to check, could the result above be because food mass differed at the start of the experiment?
lm_foodinitial <- lm(initialfoodmass~foodtype, data = food) 
anova(lm_foodinitial)

plot(lm_foodinitial) 

hist(food$initialfoodmass)
```

## SURVIVAL TRIAL ANALYSIS 

```{r survival analysis, echo=TRUE}
# perform analysis
cox_survival<- coxph(formula = Surv(time, status) ~ diet.num, data = corncox)
cox_survival
```

```{r}
res.cox.fit <- survfit(Surv(time, status) ~ diet, data = corncox)
ggsurvplot(res.cox.fit)
cox_z_survival<-cox.zph(cox_survival)
ggcoxzph(cox_z_survival)
```

```{r}
# Plot for cox regression (Thanks, Connor!)

  # export just the plot so that it can be more easily worked into multi-panel settings
test= ggsurvplot(res.cox.fit)
gg.surv= test$plot

gg.surv= gg.surv +
  xlab("Day") +
  scale_color_manual(values=c("#E69F00","#7fcdbb", "#56B4E9"))+
  theme_classic(base_size = 18)+
  theme(legend.position = "none") 

survival.plot= ggsurvplot(res.cox.fit,
                          palette= c("#E69F00","#7fcdbb", "#56B4E9"),
                          legend.title = expression(paste(italic("Ophiocordyceps"), " sp. Diet Treatment")),
                          legend.labs = c("High","Low", "Control"),
                          legend= c(0.8, 0.8)) +
  xlab("Days") 

gg.surv
```
